---
title: "Stat 240 Project Proposal"
output: html_document
author: Will Deley, Anish Mantri, Alexander Teran
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Data

```{r}
library(tidyverse)
```


## A brief description of how the group will communicate and work together.

- 1.When during the week can the group meet?
 
  Wednesday and Friday at 4 after group members classes are over. 
  
- 2.What platforms will the group use to communicate and share documents?
 
 The group has already created a github to share documents and the group has also exchanged phone numbers for mode of communication


## The question or questions of interest that motivate the planned study.
-  "How does shot location for each team effect success rate in the 2024-2025 NBA playoffs?” Is the question we will be trying to answer
- This question investigates whether a team’s shooting strategy — specifically, how often shots are taken from different areas on the court influences overall success, measured by shooting efficiency or playoff performance.


## The source(s) of data.
- Source of data: https://www.nba.com/stats
- Full link: https://www.nba.com/stats/players/shooting?DistanceRange=By%20Zone&PerMode=PerGame&Season=2024-25&SeasonType=Playoffs

```{r}
nba_24_25_playoff_shooting_data <- read_csv("./nba_24-25_playoff_shooting_by_zone.csv")
nba_24_25_playoff_shooting_data
```



## A description of the data. Include in this description answers to these questions:
- Who collected the data: The National Basketball Association 
- Data is collected through cameras(Second Spectrum) throughout the NBA playoffs and box scores written down by statisticians during games
- Data was scraped from the offical NBA stats page into a csv file using a python script
-  Key Variables : Player/Team || fg_percentage || three_pt_percentage || shot_location || attempts 

**Relation to Research Question:**  
The variables shot_location and three_pt_percentage capture where shots are taken and how effective they are for each player, while team_id allows us to aggregate across players to measure team strategy and success.

**What a Single Row Represents:**  
A single players average performance per game in each of the variabel stats in the 2024-25 playoffs. Including data such as age, team, and fg % from different shot locations.

**Larger Population:**  
This dataset can be thought of as a sample from the broader population of all NBA players and teams across seasons.




## Note: if you get the data from a data repository such as Kaggle, you will need to dig into supplementary information to answer these questions.


## Include a description of a graph which will show how the data informs the primary question of interest.
We plan to create 2 **scatter plots** where each point represents a team in the 2024–2025 NBA playoffs:
- **x-axis:** Proportion of shots taken from 3-point range (or near the basket)  
- **y-axis:** Team success rate (overall field goal percentage or win rate)
- **Color:** May represent playoff round or average player efficiency

This visualization will help reveal whether teams that take more close-range shots tend to have higher success rates.


## Include a description of a possible model for the data and associated methods of inference or prediction.
We plan to use **correlation analysis** and possibly a **simple linear regression model** to evaluate whether there is a significant relationship between shot selection and team success rate.

**Example model:**
\[
\text{Team Success Rate} = \beta_0 + \beta_1(\text{Proportion of Close Shots}) + \epsilon
\]

- **Dependent variable:** Team success rate (e.g., FG% or win rate)  
- **Independent variable:** Proportion of shots taken near the basket  

We will test whether the slope parameter (\(\beta_1\)) is significantly greater than zero, suggesting that shot location strategy impacts team performance.



